

## File names


```{r load libraries}
library(tidyverse)
library(readxl)

setwd(r"(C:\Users\anlr0006\repos-win\DOC_catchments)")

####################### Input ##################################

file_shp <- r"(\\storage.slu.se\Home$\anlr0006\My Documents\04_Projects\11_Lakes\01_data\02_raw_data\catchments\corrected_merged\corrected_merged.shp)" # replace with your location of catchment files

# Chemistry
file_chemistry <- r"(C:\Users\anlr0006\repos-win\DOC_catchments\test_results\chemistry\chem_combined.csv)" # change to your chem combined file, the output of get_water_chem.r
file_chemistry_processed <-r"(\\storage.slu.se\Home$\anlr0006\My Documents\04_Projects\11_Lakes\01_data\03_processed_data\chemistry\water_chem_all_2025.csv)"   # replace with your location of your processed water chem file (will be created in below chunk)

# Climate
folder_climate =  "\\\\storage.slu.se\\Home$\\anlr0006\\My Documents\\04_Projects\\11_Lakes\\01_data\\03_processed_data\\inter_climate"
file_climate_processed <- r"(\\storage.slu.se\Home$\anlr0006\My Documents\04_Projects\11_Lakes\03_publications\SND\data\etc\climate_merged.csv)"

# Spatial Mapped Data
NMD_file1 <- r"(\\storage.slu.se\Home$\anlr0006\My Documents\04_Projects\11_Lakes\01_data\02_raw_data\NMD\all_outputs_trend_and_survey_lakes.xlsx)"
NMD_file2 <- r"(\\storage.slu.se\Home$\anlr0006\My Documents\04_Projects\11_Lakes\01_data\01_received\all_outputs_LakeSurvey_aro251001.xlsx)"
file_HK <- r"(\\storage.slu.se\Home$\anlr0006\My Documents\04_Projects\11_Lakes\03_publications\SND\data\etc\high_coast_ecoregions_merged.csv)"

# Discharge
file_discharge.omdrev <- r"(\\storage.slu.se\Home$\anlr0006\My Documents\04_Projects\11_Lakes\03_publications\SND\data\etc\omdrev_discharge.csv)"
file_discharge.trend <- r"(\\storage.slu.se\Home$\anlr0006\My Documents\04_Projects\11_Lakes\03_publications\SND\data\etc\trendlakes_discharge.csv)"
file_discharge.combined <- r"(\\storage.slu.se\Home$\anlr0006\My Documents\04_Projects\11_Lakes\03_publications\SND\data\etc\discharge_merged.csv)"

# Deposition EMEP
emep_file.1 <- r"(\\storage.slu.se\Home$\anlr0006\My Documents\04_Projects\11_Lakes\03_publications\SND\data\etc\emep_trend.csv)"
emep_file.2 <- r"(\\storage.slu.se\Home$\anlr0006\My Documents\04_Projects\11_Lakes\03_publications\SND\data\etc\\emep_omdrev.csv)"

# NDVI

folder_NDVI <-  r"(\\storage.slu.se\Home$\anlr0006\My Documents\04_Projects\11_Lakes\01_data\02_raw_data\NDVI)"
file_NDVI <- r"(\\storage.slu.se\Home$\anlr0006\My Documents\04_Projects\11_Lakes\03_publications\SND\data\etc\NDVI_merged.csv)"

# Peat and ditches
file_peat <- r"(\\storage.slu.se\Home$\anlr0006\My Documents\04_Projects\11_Lakes\03_publications\SND\data\etc\peat_ditches.csv)"

# Soil depth
soil_depth_file <- r"(\\storage.slu.se\Home$\anlr0006\My Documents\04_Projects\11_Lakes\03_publications\SND\data\etc\soil_depth.csv)"


############################ RESULTS ##########################################

# Chemistry Time series

file_chemistry.sls <- r"(\\storage.slu.se\Home$\anlr0006\My Documents\04_Projects\11_Lakes\03_publications\SND\data\chemistry_SLS.csv)" 

file_chemistry.trend <- r"(\\storage.slu.se\Home$\anlr0006\My Documents\04_Projects\11_Lakes\03_publications\SND\data\chemistry_trend.csv)"


# Drivers

file_drivers.sls <- r"(\\storage.slu.se\Home$\anlr0006\My Documents\04_Projects\11_Lakes\03_publications\SND\data\characteristics_sls.csv)" 

file_drivers.trend <- r"(\\storage.slu.se\Home$\anlr0006\My Documents\04_Projects\11_Lakes\03_publications\SND\data\characteristics_trend.csv)" 

```

## Functions

These functions below are used further along the script to calculate catchment charcetristics
```{r}
get_mean_or_first_pre2018 <- function(data, var, date_var = "sampling_date") {
  start_period <- filter(data, year >= 2008, year <= 2012)
  val <- mean(start_period[[var]], na.rm = TRUE)
  
  if (is.na(val)) {
    pre2018 <- data %>%
      filter(year(.data[[date_var]]) < 2018 & !is.na(.data[[var]])) %>%
      arrange(.data[[date_var]]) %>%
      slice(1)
    
    if (nrow(pre2018) == 1) {
      return(pre2018[[var]])
    } else {
      return(NA)
    }
  } else {
    return(val)
  }
}

get_variable_change <- function(df, varname) {
  # Ensure date column exists and is of Date type
  if (!"sampling_date" %in% colnames(df)) stop("Date column not found")
  
  df <- df[!is.na(df[[varname]]), ]  # Remove NA values in the variable column
  
  # Filter first observation after 2012
  df_after_2012 <- df[df$sampling_date > as.Date("2012-12-31"), ]
  first_after_2012 <- df_after_2012[order(df_after_2012$sampling_date), ][1, ]
  
  # Filter first observation before 2025
  df_before_2025 <- df[df$sampling_date < as.Date("2025-01-01"), ]
  last_before_2025 <- df_before_2025[order(df_before_2025$sampling_date, decreasing = TRUE), ][1, ]
  
  # Check if we have both observations
  if (nrow(first_after_2012) == 0 || nrow(last_before_2025) == 0) {
    return(NA)  # Not enough data to compute change
  }
  
  delta_years <- year(last_before_2025$sampling_date) - year(first_after_2012$sampling_date)
  
  delta_value <- last_before_2025[[varname]] - first_after_2012[[varname]]
  
  # Ensure delta_years is valid
  if (is.na(delta_years)) {
   # warning("delta_years is NA – returning NA")
    return(NA_real_)
  }
  
  if (delta_years == 0) {
   # warning("delta_years is zero – returning NA")
    return(NA_real_)
  }
  
  return(delta_value / delta_years)
}

theil_sen_slope <- function(x, y) {
  df <- data.frame(x = as.numeric(x), y = y)
  if (nrow(df) < 2 || all(is.na(df$y))) return(NA)
  model <- mblm(y ~ x, data = df, repeated = FALSE)
  coef(model)[2]
}

# Function to compute 0.75 * IQR / median
robust_variation <- function(x) {
  if (median(x, na.rm = TRUE) == 0) return(NA)
  0.75 * IQR(x, na.rm = TRUE) / median(x, na.rm = TRUE)
}

library(rlang)

slope_or_na <- function(column_name, data) {
  # Make sure the column exists
  if (!column_name %in% names(data)) return(NA_real_)
  
  # Filter out rows where y (the column) OR x (year) is NA
  df <- data %>%
    select(year, all_of(column_name)) %>%
    drop_na()
  
  # Only compute if there are at least 10 non-NA rows
  if (nrow(df) < 10) return(NA_real_)
  
  # Fit Theil-Sen slope
  model <- try(mblm::mblm(as.formula(paste(column_name, "~ year")), data = df), silent = TRUE)
  
  if (inherits(model, "try-error")) return(NA_real_)
  
  coef(model)[2]
}

```

## Catchments 

Aggregating all the extracted spatial and temporal drivers to the right resolution. 


```{r}
library(tidyverse)
library(sf)
shp <- sf::read_sf(file_shp)  

# This line is to drop duplicate id's this is because trend and omdrev sometimes have the same lakes. 
shp <- shp[!duplicated(shp$mvm_id), ]

shp %>% select(x_utlopp, y_utlopp, mvm_id, Name) %>% st_drop_geometry() %>% rename(x_e_koord = x_utlopp, y_n_koord = y_utlopp,  stationname = Name) %>% mutate(mvm_id = as.numeric(mvm_id)) -> coordinates

id.all <- shp$mvm_id

df_sf <- st_as_sf(coordinates, coords = c("x_e_koord", "y_n_koord"), crs = 3006)
df_wgs84 <- st_transform(df_sf, crs = 4326)

# Extract transformed coordinates (lon/lat)
coords <- st_coordinates(df_wgs84)
coordinates$lon <- coords[, 1]
coordinates$lat <- coords[, 2]

coordinates
```


# Processing and loading data

## Chemistry

This chunk converts all the chemistry data into the standard units.
As the API gives us multiple different units this ensures in the end each station uses the same standard unit for each variable.
Some additional chemistry variables are calculated and values below the detection limit are replaced with half the detection limit. 
Any station with only a single sample is removed.



```{r}
library(vroom)
library(dplyr)
library(lubridate)


# df.chem <- read.csv(file_chemistry) %>% as_tibble() #uncomment if you only have one chemistry file
# df.chem


# Comment out if you have just one chemistry file
# file_chemistry1 <- r"(\\storage.slu.se\Home$\anlr0006\My Documents\04_Projects\11_Lakes\01_data\02_raw_data\chemistry\water_chem_combined_catchments_2025.csv)" 
# file_chemistry2 <- r"(\\storage.slu.se\Home$\anlr0006\My Documents\04_Projects\11_Lakes\01_data\02_raw_data\chemistry\water_chem_combined_catchments_post_2025.csv)" 
# 
# file_chemistry3 <- r"(\\storage.slu.se\Home$\anlr0006\My Documents\04_Projects\11_Lakes\01_data\02_raw_data\chemistry\water_chem_combined_catchments_pre_2025.csv)"
# 
# df.chem <- read.csv(file_chemistry1) %>% bind_rows(read.csv(file_chemistry2)) %>% bind_rows(read.csv(file_chemistry3)) %>%
#   distinct(.keep_all = TRUE) %>% as_tibble() %>% select(-c(X))

convertible <- sapply(df.chem, function(col) {
  all(is.na(col) | !is.na(suppressWarnings(as.numeric(as.character(col)))))
})



# selecting only omdrev and trend studyName, remove this part in case you care about all the samples taken at a station, no matter what program
df.chem <- df.chem %>% mutate(sampling_date = as.Date(sampling_date))%>% filter(grepl('trend|omdre', studyName)) %>% # if there are multiple samples on the same day , pick the one closest to 0.5 
  mutate(across(-c(sampling_date,subprogramName, studyName), ~ suppressWarnings(as.numeric(as.character(.)))))
  

safe_col <- function(df, col) {
  if (col %in% names(df)) df[[col]] else rep(NA, nrow(df))
}

# safe_coalesce <- function(df, ...) {
#   cols <- list(...)
#   vals <- lapply(cols, function(col) safe_col(df, col))
#   dplyr::coalesce(!!!vals)
# }

  
df.chem <-  df.chem %>%
  mutate(
    # TOC
    toc_mg_l = coalesce(
      safe_col(., "TOC_mgl"),
      safe_col(., "TOC_mgl.C")
    ),
    toc_origin = case_when(
      !is.na(safe_col(., "TOC_mgl")) ~ "TOC_mgl",
      !is.na(safe_col(., "TOC_mgl.C")) ~ "TOC_mgl.C",
      TRUE ~ NA_character_
    )
  ) %>%
  drop_na(toc_mg_l) %>%
  group_by(mvm_id, sampling_date, studyName) %>%
  slice_min(order_by = abs(min_depth - 0.5), n = 1, with_ties = FALSE) %>%
  ungroup() %>%
  mutate(
    # Sulfate thresholds, detection limits 
    SO4_IC_mgl.SO4 = ifelse(
      safe_col(., "SO4_IC_mgl.SO4") <= 0.48,
      0.24,
      safe_col(., "SO4_IC_mgl.SO4")
    ),
    SO4_IC_mgl = ifelse(
      safe_col(., "SO4_IC_mgl") <= 0.16,
      0.08,
      safe_col(., "SO4_IC_mgl")
    ),
    SO4_mgl.S = ifelse(
      safe_col(., "SO4_mgl.S") <= 0.2,
      0.1,
      safe_col(., "SO4_mgl.S")
    ),

    # Tot N / TNb adjustments
    `Tot_N_TNb_µgl.N` = ifelse(
      (year(sampling_date) + (month(sampling_date) - 1) / 12) > (2009 + 8/12),
      safe_col(., "Tot_N_TNb_µgl.N"),
      NA
    ),
    `Tot_N_TNb_mgl.N` = ifelse(
      (year(sampling_date) + (month(sampling_date) - 1) / 12) > (2009 + 8/12),
      safe_col(., "Tot_N_TNb_mgl.N"),
      NA
    ),

    tot_n_t_nb_mg_l_n = coalesce(
      safe_col(., "Tot_N_mgl"),
      safe_col(., "Tot_N_mgl.N"),
      safe_col(., "Tot_N_µgl") / 1000,
      safe_col(., "Tot_N_µgl.N") / 1000,
      safe_col(., "Tot_N_TNb_µgl.N") / 1000,
      safe_col(., "Tot_N_TNb_µgl") / 1000,
      safe_col(., "Tot_N_TNb_mgl"),
      safe_col(., "Tot_N_TNb_mgl.N"),
      safe_col(., "Tot_N_summa_µgl") / 1000,
      safe_col(., "Tot_N_summa_µgl.N") / 1000,
      safe_col(., "Tot_N_ps_µgl.N") / 1000,
      safe_col(., "Tot_N_ps_µgl") / 1000,
      safe_col(., "Tot_N_ps_mgl"),
      safe_col(., "Tot_N_ps_mgl.N")
    ),

    # Sulfate in mol/L & origin
    so4_mol = coalesce(
      safe_col(., "SO4_IC_mekvl") / (2 * 1000),
      safe_col(., "SO4_mekvl") / (2 * 1000),
      safe_col(., "SO4_IC_mgl") / (96.0626 * 1000),
      safe_col(., "SO4_IC_mgl.SO4") / (96.0626 * 1000),
      safe_col(., "SO4_mgl.SO4") / (96.0626 * 1000),
      safe_col(., "SO4_mgl.S") / (32.06 * 1000),
      safe_col(., "SO4_mgl") / (96.0626 * 1000)
    ),
    so4_origin = case_when(
      !is.na(safe_col(., "SO4_IC_mekvl")) ~ "SO4_IC_mekvl",
      !is.na(safe_col(., "SO4_mekvl")) ~ "SO4_mekvl",
      !is.na(safe_col(., "SO4_IC_mgl")) ~ "SO4_IC_mgl",
      !is.na(safe_col(., "SO4_IC_mgl.SO4")) ~ "SO4_IC_mgl.SO4",
      !is.na(safe_col(., "SO4_mgl.SO4")) ~ "SO4_mgl.SO4",
      !is.na(safe_col(., "SO4_mgl.S")) ~ "SO4_mgl.S",
      !is.na(safe_col(., "SO4_mgl")) ~ "SO4_mgl",
      TRUE ~ NA_character_
    ),

    # Iron (Fe)
    fe_mg_l = coalesce(
      safe_col(., "Fe_F_mgl"),
      safe_col(., "Fe_F_µgl") / 1000,
      safe_col(., "Fe_mgl"),
      safe_col(., "Fe_µgl") / 1000
    ),
    fe_origin = case_when(
      !is.na(safe_col(., "Fe_F_mgl")) ~ "Fe_F_mgl",
      !is.na(safe_col(., "Fe_F_µgl")) ~ "Fe_F_µgl",
      !is.na(safe_col(., "Fe_mgl")) ~ "Fe_mgl",
      !is.na(safe_col(., "Fe_µgl")) ~ "Fe_µgl",
      TRUE ~ NA_character_
    ),

    # Silicon (Si)
    si_mg_l = coalesce(
      safe_col(., "Si_F_mgl"),
      safe_col(., "Si_µgl") / 1000,
      safe_col(., "Si_mgl")
    ),
    si_origin = case_when(
      !is.na(safe_col(., "Si_F_mgl")) ~ "Si_F_mgl",
      !is.na(safe_col(., "Si_µgl")) ~ "Si_µgl",
      !is.na(safe_col(., "Si_mgl")) ~ "Si_mgl",
      TRUE ~ NA_character_
    ),

    # Aluminium (Al)
    al_mg_l = coalesce(
      safe_col(., "Al_F_µgl") / 1000,
      safe_col(., "Al_ICP_µgl") / 1000,
      safe_col(., "Al_ICPAES_µgl") / 1000,
      safe_col(., "Al_s_µgl") / 1000,
      safe_col(., "Al_µgl") / 1000
    ),
    al_origin = case_when(
      !is.na(safe_col(., "Al_F_µgl")) ~ "Al_F_µgl",
      !is.na(safe_col(., "Al_ICP_µgl")) ~ "Al_ICP_µgl",
      !is.na(safe_col(., "Al_ICPAES_µgl")) ~ "Al_ICPAES_µgl",
      !is.na(safe_col(., "Al_s_µgl")) ~ "Al_s_µgl",
      !is.na(safe_col(., "Al_µgl")) ~ "Al_µgl",
      TRUE ~ NA_character_
    ),

    # Ammonium (NH4‑N)
    nh4_n_mg_l_n = coalesce(
      safe_col(., "NH4_N_mgl"),
      safe_col(., "NH4_N_µgl") / 1000,
      safe_col(., "NH4_N_µgl.N") / 1000
    ),
    nh4_n_mg_l_n_origin = case_when(
      !is.na(safe_col(., "NH4_N_mgl")) ~ "NH4_N_mgl",
      !is.na(safe_col(., "NH4_N_µgl")) ~ "NH4_N_µgl",
      !is.na(safe_col(., "NH4_N_µgl.N")) ~ "NH4_N_µgl.N",
      TRUE ~ NA_character_
    ),

    # Calcium (Ca²⁺)
    ca_mg_l = coalesce(
      safe_col(., "Ca_F_mgl"),
      (safe_col(., "Ca_mekvl") / 2) * 40.48,
      safe_col(., "Ca_mgl")
    ),
    ca_origin = case_when(
      !is.na(safe_col(., "Ca_F_mgl")) ~ "Ca_F_mgl",
      !is.na(safe_col(., "Ca_mekvl")) ~ "Ca_mekvl",
      !is.na(safe_col(., "Ca_mgl")) ~ "Ca_mgl",
      TRUE ~ NA_character_
    ),

    # Potassium (K)
    k_mg_l = coalesce(
      safe_col(., "K_F_mgl"),
      safe_col(., "K_mekvl") * 39.1,
      safe_col(., "K_mgl")
    ),
    k_origin = case_when(
      !is.na(safe_col(., "K_F_mgl")) ~ "K_F_mgl",
      !is.na(safe_col(., "K_mekvl")) ~ "K_mekvl",
      !is.na(safe_col(., "K_mgl")) ~ "K_mgl",
      TRUE ~ NA_character_
    ),

    # Chloride (Cl⁻)
    cl_mg_l = coalesce(
      safe_col(., "Cl_mekvl") * 35.45,
      safe_col(., "Cl_mgl")
    ),
    cl_origin = case_when(
      !is.na(safe_col(., "Cl_mekvl")) ~ "Cl_mekvl",
      !is.na(safe_col(., "Cl_mgl")) ~ "Cl_mgl",
      TRUE ~ NA_character_
    ),

    # NO2 + NO3 as N
    no2_no3_n_mg_l_n = coalesce(
      safe_col(., "NO2_NO3_N_µgl") / 1000,
      safe_col(., "NO2_NO3_N_mgl"),
      safe_col(., "NO2_NO3_N_µgl.N") / 1000,
      safe_col(., "NO3_N_µgl") / 1000,
      safe_col(., "NO3_N_mgl")
    ),
    no2_no3_n_mg_l_n_origin = case_when(
      !is.na(safe_col(., "NO2_NO3_N_µgl")) ~ "NO2_NO3_N_µgl",
      !is.na(safe_col(., "NO2_NO3_N_mgl")) ~ "NO2_NO3_N_mgl",
      !is.na(safe_col(., "NO2_NO3_N_µgl.N")) ~ "NO2_NO3_N_µgl.N",
      !is.na(safe_col(., "NO3_N_µgl")) ~ "NO3_N_µgl",
      !is.na(safe_col(., "NO3_N_mgl")) ~ "NO3_N_mgl",
      TRUE ~ NA_character_
    ),

    # Fluoride (F)
    f_mg_l = safe_col(., "Fluorid_mgl"),
    f_origin = case_when(
      !is.na(safe_col(., "Fluorid_mgl")) ~ "Fluorid_mgl",
      TRUE ~ NA_character_
    ),

    # Alkalinity / Acidity
    alk_acid_mmol = case_when(
      !is.na(safe_col(., "Alk_Acid_mekvl")) ~ safe_col(., "Alk_Acid_mekvl"),
      !is.na(safe_col(., "Alk_Acid_mmoll")) ~ safe_col(., "Alk_Acid_mmoll"),
      (!is.na(safe_col(., "Acid_mekvl")) & safe_col(., "Acid_mekvl") != 0) ~ -safe_col(., "Acid_mekvl"),
      !is.na(safe_col(., "Alk_mg.HCO3l")) ~ safe_col(., "Alk_mg.HCO3l") / 61.02,
      !is.na(safe_col(., "Alk_mekvl")) ~ safe_col(., "Alk_mekvl"),
      !is.na(safe_col(., "Alk_mmoll")) ~ safe_col(., "Alk_mmoll"),
      TRUE ~ NA_real_
    ),
    alk_acid_origin = case_when(
      !is.na(safe_col(., "Alk_Acid_mekvl")) ~ "Alk_Acid_mekvl",
      !is.na(safe_col(., "Alk_Acid_mmoll")) ~ "Alk_Acid_mmoll",
      (!is.na(safe_col(., "Acid_mekvl")) & safe_col(., "Acid_mekvl") != 0) ~ "Acid_mekvl",
      !is.na(safe_col(., "Alk_mg.HCO3l")) ~ "Alk_mg.HCO3l",
      !is.na(safe_col(., "Alk_mekvl")) ~ "Alk_mekvl",
      !is.na(safe_col(., "Alk_mmoll")) ~ "Alk_mmoll",
      TRUE ~ NA_character_
    ),

    # Sodium (Na)
    na_mg_l = coalesce(
      safe_col(., "Na_F_mgl"),
      safe_col(., "Na_mekvl") * 22.99,
      safe_col(., "Na_mgl")
    ),
    na_origin = case_when(
      !is.na(safe_col(., "Na_F_mgl")) ~ "Na_F_mgl",
      !is.na(safe_col(., "Na_mekvl")) ~ "Na_mekvl",
      !is.na(safe_col(., "Na_mgl")) ~ "Na_mgl",
      TRUE ~ NA_character_
    ),

    # Manganese (Mn)
    mn_mg_l = coalesce(
      safe_col(., "Mn_F_µgl") / 1000,
      safe_col(., "Mn_µgl") / 1000,
      safe_col(., "Mn_mgl")
    ),
    mn_origin = case_when(
      !is.na(safe_col(., "Mn_F_µgl")) ~ "Mn_F_µgl",
      !is.na(safe_col(., "Mn_µgl")) ~ "Mn_µgl",
      !is.na(safe_col(., "Mn_mgl")) ~ "Mn_mgl",
      TRUE ~ NA_character_
    ),

    # Zinc (Zn)
    zn_mg_l = coalesce(
      safe_col(., "Zn_F_mgl"),
      safe_col(., "Zn_µgl") / 1000,
      safe_col(., "Zn_mgl")
    ),
    zn_origin = case_when(
      !is.na(safe_col(., "Zn_F_mgl")) ~ "Zn_F_mgl",
      !is.na(safe_col(., "Zn_µgl")) ~ "Zn_µgl",
      !is.na(safe_col(., "Zn_mgl")) ~ "Zn_mgl",
      TRUE ~ NA_character_
    ),

    # Copper (Cu)
    cu_mg_l = coalesce(
      safe_col(., "Cu_F_mgl"),
      safe_col(., "Cu_µgl") / 1000,
      safe_col(., "Cu_mgl")
    ),
    cu_origin = case_when(
      !is.na(safe_col(., "Cu_F_mgl")) ~ "Cu_F_mgl",
      !is.na(safe_col(., "Cu_µgl")) ~ "Cu_µgl",
      !is.na(safe_col(., "Cu_mgl")) ~ "Cu_mgl",
      TRUE ~ NA_character_
    ),

    # Vanadium (V)
    v_mg_l = coalesce(
      safe_col(., "V_F_µgl") / 1000,
      safe_col(., "V_µgl") / 1000,
      safe_col(., "V_mgl")
    ),
    v_origin = case_when(
      !is.na(safe_col(., "V_F_µgl")) ~ "V_F_µgl",
      !is.na(safe_col(., "V_µgl")) ~ "V_µgl",
      !is.na(safe_col(., "V_mgl")) ~ "V_mgl",
      TRUE ~ NA_character_
    ),

    # Nickel (Ni)
    ni_mg_l = coalesce(
      safe_col(., "Ni_F_µgl") / 1000,
      safe_col(., "Ni_µgl") / 1000,
      safe_col(., "Ni_mgl")
    ),
    ni_origin = case_when(
      !is.na(safe_col(., "Ni_F_µgl")) ~ "Ni_F_µgl",
      !is.na(safe_col(., "Ni_µgl")) ~ "Ni_µgl",
      !is.na(safe_col(., "Ni_mgl")) ~ "Ni_mgl",
      TRUE ~ NA_character_
    ),

    # Lead (Pb)
    pb_mg_l = coalesce(
      safe_col(., "Pb_F_µgl") / 1000,
      safe_col(., "Pb_µgl") / 1000,
      safe_col(., "Pb_mgl")
    ),
    pb_origin = case_when(
      !is.na(safe_col(., "Pb_F_µgl")) ~ "Pb_F_µgl",
      !is.na(safe_col(., "Pb_µgl")) ~ "Pb_µgl",
      !is.na(safe_col(., "Pb_mgl")) ~ "Pb_mgl",
      TRUE ~ NA_character_
    ),

    # Electrical Conductivity (to 20 °C)
    ec20_µS_cm = coalesce(
      safe_col(., "Kond_20_µScm"),
      safe_col(., "Kond_25_µScm") * 0.90,
      safe_col(., "Kond_20_mSm") * 10,
      safe_col(., "Kond_25_mSm") * 10 * 0.90,
      safe_col(., "Kond_mSm") * 10 * 0.90
    ),
    ec20_origin = case_when(
      !is.na(safe_col(., "Kond_20_µScm")) ~ "Kond_20_µScm",
      !is.na(safe_col(., "Kond_25_µScm")) ~ "Kond_25_µScm",
      !is.na(safe_col(., "Kond_20_mSm")) ~ "Kond_20_mSm",
      !is.na(safe_col(., "Kond_25_mSm")) ~ "Kond_25_mSm",
      !is.na(safe_col(., "Kond_mSm")) ~ "Kond_mSm",
      TRUE ~ NA_character_
    ),

    # Absorbance / Fluorescence
    abs_f_420_5cm = safe_col(., "Abs_F420_5cm"),
    
  # Total Phosphorus
  tot_p_mg_l_p = coalesce(
    safe_col(., "Tot_P_µgl") / 1000,
    safe_col(., "Tot_P_µgl.P") / 1000,
    safe_col(., "Tot_P_mgl"),
    safe_col(., "Tot_P_mgl.P"),
    safe_col(., "Tot_P_F_µgl") / 1000,
    safe_col(., "Tot_P_F_µgl.P") / 1000,
    safe_col(., "Tot_P_F_mgl.P")
  ),
  
  tot_p_origin = case_when(
    !is.na(safe_col(., "Tot_P_µgl")) ~ "Tot_P_µgl",
    !is.na(safe_col(., "Tot_P_µgl.P")) ~ "Tot_P_µgl.P",
    !is.na(safe_col(., "Tot_P_mgl")) ~ "Tot_P_mgl",
    !is.na(safe_col(., "Tot_P_mgl.P")) ~ "Tot_P_mgl.P",
    !is.na(safe_col(., "Tot_P_F_µgl")) ~ "Tot_P_F_µgl",
    !is.na(safe_col(., "Tot_P_F_µgl.P")) ~ "Tot_P_F_µgl.P",
    !is.na(safe_col(., "Tot_P_F_mgl.P")) ~ "Tot_P_F_mgl.P",
    TRUE ~ NA_character_
  ),
  ca_mg_l = coalesce(
    safe_col(., "Ca_F_mgl"),                # already mg/L
    safe_col(., "Ca_mekvl") /2 *40.48,        # meq/L → mg/L (40.08 g/mol ÷ 2 eq/mol = 20.04 mg/meq)
    safe_col(., "Ca_mgl")                   # already mg/L
  ),
  
  ca_origin = case_when(
    !is.na(safe_col(., "Ca_F_mgl")) ~ "Ca_F_mgl",
    !is.na(safe_col(., "Ca_mekvl")) ~ "Ca_mekvl",
    !is.na(safe_col(., "Ca_mgl")) ~ "Ca_mgl",
    TRUE ~ NA_character_
  ),
  mg_mg_l = coalesce(
    safe_col(., "Mg_F_mgl"),                 # already mg/L
    safe_col(., "Mg_mekvl") /2 *24.305,       # meq/L → mg/L (24.305 g/mol ÷ 2 eq/mol = 12.1525 mg/meq)
    safe_col(., "Mg_mgl")                    # already mg/L
  ),
  
  mg_origin = case_when(
    !is.na(safe_col(., "Mg_F_mgl")) ~ "Mg_F_mgl",
    !is.na(safe_col(., "Mg_mekvl")) ~ "Mg_mekvl",
    !is.na(safe_col(., "Mg_mgl")) ~ "Mg_mgl",
    TRUE ~ NA_character_
  ),
  na_mg_l = coalesce(
    safe_col(., "Na_F_mgl"),            # already mg/L
    safe_col(., "Na_mekvl") * 22.99,    # meq/L → mg/L (Na⁺: 22.99 g/mol ÷ 1 eq/mol)
    safe_col(., "Na_mgl")               # already mg/L
  ),
  
  na_origin = case_when(
    !is.na(safe_col(., "Na_F_mgl")) ~ "Na_F_mgl",
    !is.na(safe_col(., "Na_mekvl")) ~ "Na_mekvl",
    !is.na(safe_col(., "Na_mgl")) ~ "Na_mgl",
    TRUE ~ NA_character_
  )
  )

df.chem %>% select(mvm_id) %>% mutate(mvm_id = as.numeric(mvm_id)) %>% unique() %>% summarize(count.unique=n())%>% pull() -> n_stations_pre_filtering

print(paste("We downloaded chemistry data from",n_stations_pre_filtering, "stations that are part of trendlakes and the swedish lake survey (SLS), the rest had data only before 1980."))


# this is a set of duplicate stations, where a single station has multiple ids
id_redirect <- c(
  "2638"  = "2021",
  "41634" = "40311",
  "41534" = "40352"
)

safe_numeric <- function(df, col) {
  if (col %in% names(df)) as.numeric(df[[col]]) else rep(NA_real_, nrow(df))
}


# setting common detection limits, this is to ensure the changes in detection limits don't result in a 'fake' trend.
# Values below the detection limits are replaced with half the detection limit.
df.chem <- df.chem %>%
  mutate(
    si_mg_l = case_when(safe_numeric(., "si_mg_l") <= 0.5 ~ 0.25,
                        TRUE ~ safe_numeric(., "si_mg_l")),
    cl_mg_l = case_when(safe_numeric(., "cl_mg_l") <= 0.25 ~ 0.125,
                        TRUE ~ safe_numeric(., "cl_mg_l")),
    zn_mg_l = case_when(safe_numeric(., "zn_mg_l") <= 0.0005 ~ 0.00025,
                        TRUE ~ safe_numeric(., "zn_mg_l")),
    v_mg_l = case_when(safe_numeric(., "v_mg_l") <= 0.00003 ~ 0.000015,
                       TRUE ~ safe_numeric(., "v_mg_l")),
    pb_mg_l = case_when(safe_numeric(., "pb_mg_l") <= 0.00002 ~ 0.00001,
                        TRUE ~ safe_numeric(., "pb_mg_l")),
    mn_mg_l = case_when(safe_numeric(., "mn_mg_l") <= 0.0005 ~ 0.00025,
                        TRUE ~ safe_numeric(., "mn_mg_l")),
    ni_mg_l = case_when(safe_numeric(., "ni_mg_l") <= 0.00005 ~ 0.000025,
                        TRUE ~ safe_numeric(., "ni_mg_l")),
    fe_mg_l = case_when(safe_numeric(., "fe_mg_l") <= 0.01 ~ 0.005,
                        TRUE ~ safe_numeric(., "fe_mg_l")),
    f_mg_l = case_when(safe_numeric(., "f_mg_l") <= 0.05 ~ 0.025,
                       TRUE ~ safe_numeric(., "f_mg_l")),
    cu_mg_l = case_when(safe_numeric(., "cu_mg_l") < 0.00004 ~ 0.00002,
                        TRUE ~ safe_numeric(., "cu_mg_l")),
    cr_mg_l = case_when(safe_numeric(., "cr_mg_l") <= 0.00005 ~ 0.000025,
                        TRUE ~ safe_numeric(., "cr_mg_l")),
    co_mg_l = case_when(safe_numeric(., "co_mg_l") <= 0.00001 ~ 0.000005,
                        TRUE ~ safe_numeric(., "co_mg_l")),
    cd_mg_l = case_when(safe_numeric(., "cd_mg_l") <= 0.000006 ~ 0.000003,
                        TRUE ~ safe_numeric(., "cd_mg_l")),
    as_mg_l = case_when(safe_numeric(., "as_mg_l") <= 0.000003 ~ 0.0000015,
                        TRUE ~ safe_numeric(., "as_mg_l")),
    k_mg_l = case_when(safe_numeric(., "k_mg_l") <= 0.04 ~ 0.02,
                       TRUE ~ safe_numeric(., "k_mg_l")),
    al_mg_l = case_when(safe_numeric(., "al_mg_l") < 0.005 ~ 0.0025,
                        TRUE ~ safe_numeric(., "al_mg_l")),
    nh4_n_mg_l_n = case_when(safe_numeric(., "nh4_n_mg_l_n") <= 0.004 ~ 0.002,
                             TRUE ~ safe_numeric(., "nh4_n_mg_l_n")),
    no2_no3_n_mg_l_n = case_when(safe_numeric(., "no2_no3_n_mg_l_n") <= 0.003 ~ 0.0015,
                                 TRUE ~ safe_numeric(., "no2_no3_n_mg_l_n")),
    toc_mg_l_c = case_when(safe_numeric(., "toc_mg_l") <= 0.05 ~ 0.025,
                           TRUE ~ safe_numeric(., "toc_mg_l")),
    totn = case_when(safe_numeric(., "tot_n_t_nb_mg_l_n") <= 0.050 ~ 0.025,
                     TRUE ~ safe_numeric(., "tot_n_t_nb_mg_l_n")),
    totp = case_when(safe_numeric(., "tot_p_mg_l_p") <= 0.001 ~ 0.0005,
                     TRUE ~ safe_numeric(., "tot_p_mg_l_p")),
    absf = case_when(safe_numeric(., "abs_f_420_5cm") <= 0.01 ~ 0.005,
                     TRUE ~ safe_numeric(., "abs_f_420_5cm")),

    # Derived values
    DIN_mg_l_n = safe_numeric(., "nh4_n_mg_l_n") + safe_numeric(., "no2_no3_n_mg_l_n"),
    BC_mekv_l = (
      safe_numeric(., "ca_mg_l") / 40.078 * 2 +
      safe_numeric(., "mg_mg_l") / 24.305 * 2 +
      safe_numeric(., "na_mg_l") / 22.99 +
      safe_numeric(., "k_mg_l") / 39.01
    ),
    ANC_mekv_l = (
      safe_numeric(., "ca_mg_l") / 40.078 * 2 +
      safe_numeric(., "mg_mg_l") / 24.305 * 2 +
      safe_numeric(., "na_mg_l") / 22.99 +
      safe_numeric(., "k_mg_l") / 39.01
    ) - (
      safe_numeric(., "cl_mg_l") / 35.45 +
      safe_numeric(., "so4_mol") * 2 * 1000 +
      safe_numeric(., "no2_no3_n_mg_l_n") / 14
    ),

    year = year(sampling_date),
    pH = safe_numeric(., "pH_")
  ) %>%
  select(
    si_mg_l, cl_mg_l, zn_mg_l, v_mg_l, pb_mg_l, mn_mg_l, ni_mg_l,
    fe_mg_l, f_mg_l, cu_mg_l, cr_mg_l, co_mg_l, cd_mg_l, as_mg_l,
    k_mg_l, al_mg_l, nh4_n_mg_l_n, no2_no3_n_mg_l_n, toc_mg_l_c, ca_mg_l, na_mg_l, mg_mg_l,
    totn, totp, absf, DIN_mg_l_n, BC_mekv_l, ANC_mekv_l, pH, sampling_date,
    sample_id, min_depth, max_depth, subprogramName, studyName, mvm_id,
    alk_acid_mmol, ec20_µS_cm, so4_mol, year
  )

# stations with less than 1 sample 2013-2024are removed
stations_with_omdre <- df.chem %>%
  filter(grepl("omdre", studyName, ignore.case = TRUE)) %>%   # only "omdre" studies
  mutate(mvp_id = recode(as.character(mvm_id), !!!id_redirect)) %>%
  filter(lubridate::year(sampling_date) >= 2013,
         lubridate::year(sampling_date) <= 2024) %>%
  count(mvp_id) %>%
  filter(n > 1) %>%
  pull(mvp_id)

stations_with_trend <- df.chem %>%
  filter(grepl("trend", studyName, ignore.case = TRUE)) %>%   # only "trend" studies
  mutate(mvp_id = recode(as.character(mvm_id), !!!id_redirect)) %>%
  filter(lubridate::year(sampling_date) >= 2013,
         lubridate::year(sampling_date) <= 2024) %>%
  count(mvp_id) %>%
  filter(n > 1) %>%
  pull(mvp_id)

stations_with_enough_data <- c(108,stations_with_trend, stations_with_omdre)


df.chem <- df.chem %>%
  mutate(mvp_id = recode(as.character(mvm_id), !!!id_redirect)) %>%
  filter(mvp_id %in% stations_with_enough_data)

df.chem

# rm(df.chem)

```
Save the preprocessed chemistry file / files. 
```{r}
## This data contains only the trend and omdrev data from these stations but it still contains all the seasons

df.chem %>% left_join(coordinates %>% mutate(mvm_id = as.numeric(mvm_id)))%>% write.csv(file_chemistry_processed)


df.chem  %>% filter( grepl('trend', studyName))  %>% group_by(mvm_id) %>% filter(n() > 1) %>% ungroup()-> df.chem.trend

df.chem  %>% filter( grepl('omdre', studyName) ) %>% group_by(mvm_id) %>% filter(n() > 1) -> df.chem.sls

df.chem.sls %>% left_join(coordinates%>% mutate(mvm_id = as.numeric(mvm_id)))%>% write.csv(file_chemistry.sls)

df.chem.trend %>% left_join(coordinates%>% mutate(mvm_id = as.numeric(mvm_id)))%>% write.csv(file_chemistry.trend)


df.chem

```

## Climate Data

Trend lakes 
```{r}
library(tidyverse)
library(mblm)


file_list <- list.files(path = folder_climate, pattern = "*\\.csv$", full.names = FALSE)

file_ids <- gsub(
  "(.*)\\.csv$", "\\1", file_list)

missing_ids <- setdiff(id.all, file_ids)
```

```{r}
library(dplyr)
library(purrr)
library(mblm)
library(readr)  # For faster reading
library(lubridate)


# --- Safely process each ID ---
climate_results <- map_dfr(id.all, function(id) {
  tryCatch({
    csv.path <- paste0(folder_climate,"\\" , id, ".csv")
    climate <- read_csv(csv.path, show_col_types = FALSE) %>%
      mutate(
        sampling_date = as.Date(date),
        year = year(sampling_date)
      ) %>%
      rename(
        snow_threshold = band_data,
        temp = tas
      )

    # --- Start period ---
    start_period <- filter(climate, year >= 2008, year <= 2012)

    temp_start <- mean(start_period$temp, na.rm = TRUE)

    pr_start <- start_period %>%
      group_by(year) %>%
      summarise(total_pr = sum(pr, na.rm = TRUE), .groups = "drop") %>%
      summarise(mean_total_pr = mean(total_pr, na.rm = TRUE)) %>%
      pull(mean_total_pr)

    pr_snow_start <- start_period %>%
      group_by(year) %>%
      summarise(total_pr_snow = sum(pr_snow, na.rm = TRUE), .groups = "drop") %>%
      summarise(mean_total_pr_snow = mean(total_pr_snow, na.rm = TRUE)) %>%
      pull(mean_total_pr_snow)

    # --- Change period ---
    annual_stats <- filter(climate, year >= 2013, year <= 2024) %>%
      group_by(year) %>%
      summarise(
        mean_temp = mean(temp, na.rm = TRUE),
        total_pr = sum(pr, na.rm = TRUE),
        total_pr_snow = sum(pr_snow, na.rm = TRUE),
        .groups = "drop"
      )

    temp_change <- if (nrow(annual_stats) >= 2) coef(mblm(mean_temp ~ year, data = annual_stats))[2] else NA
    pr_change <- if (nrow(annual_stats) >= 2) coef(mblm(total_pr ~ year, data = annual_stats))[2] else NA
    pr_snow_change <- if (nrow(annual_stats) >= 2) coef(mblm(total_pr_snow ~ year, data = annual_stats))[2] else NA

    temp_var <- robust_variation(annual_stats$mean_temp)
    pr_var <- robust_variation(annual_stats$total_pr)
    pr_snow_var <- robust_variation(annual_stats$total_pr_snow)

    tibble(
      mvm_id = id,
      temp_start = temp_start,
      temp_change = temp_change,
      temp_variation = temp_var,
      pr_start = pr_start,
      pr_change = pr_change,
      pr_variation = pr_var,
      pr_snow_start = pr_snow_start,
      pr_snow_change = pr_snow_change,
      pr_snow_variation = pr_snow_var
    )
  }, error = function(e) {
    message(paste("Error processing ID:", id, "-", e$message))
    return(NULL)  # Skip this ID
  })
})

climate_results%>% write_csv(file_climate_processed)

```

```{r}
library(vtable)
climate.all <-  read_csv(file_climate_processed) # %>% rename(mvm_id = id)
climate.all %>% sumtable()
```


# Elevation


```{r}
library(readxl)

general1 <- read_excel(NMD_file1, sheet = "aro_survey_and_trend_lakes_Merg") %>% as_tibble() %>% transmute(mvm_id = as.numeric(mvmid), 
                      lake_area_m2 = na_if(lake_area_, 0),
                      elevation_m = elev_utl,
                      area_m2 = Shape_Area) # %>% select(-all_of(c("Unnamed: 0","...1" , "area_ARO_m2"    , "area_ARO_km2"  )))

general2 <- read_excel(NMD_file2, sheet = "aro_omdrev_AnnaL251001") %>% as_tibble() %>% transmute(mvm_id = as.numeric(mvmid), 
                      lake_area_m2 = na_if(lake_area_, 0),
                      elevation_m = elev_utl,
                      area_m2 = Shape_Area) # %>% select(-all_of(c("Unnamed: 0","...1" , "area_ARO_m2"    , "area_ARO_km2"  )))

general <- general1 %>% filter(!mvm_id %in% general2$mvm_id) %>%  bind_rows(general2)


general
```

## NMD

```{r}
library(readxl)

NMD1 <- read_excel(NMD_file1, sheet = "nmd") %>% as_tibble() %>% select(-all_of(c("Unnamed: 0","...1" , "area_ARO_m2"    , "area_ARO_km2"  ))) %>% transmute(mvm_id = mvmid,
                  `forest_wetland` = rowSums(select(., contains("på våtmark")), na.rm = TRUE),
                  open_wetland = `Öppen våtmark`,
                  open_water = `Sjö och vattendrag` + `Hav`,
                  forest = rowSums(select(., contains("skog")), na.rm = TRUE),
                  semi_urban = rowSums(select(., contains("ploaterad mark")), na.rm = TRUE),
                  agriculture = `Åkermark`
                  )
NMD2 <- read_excel(NMD_file2, sheet = "nmd") %>% as_tibble() %>% select(-all_of(c("...1" , "area_ARO_m2"    , "area_ARO_km2"  ))) %>% transmute(mvm_id = mvmid,
                  `forest_wetland` = rowSums(select(., contains("på våtmark")), na.rm = TRUE),
                  open_wetland = `Öppen våtmark`,
                  open_water = `Sjö och vattendrag` + `Hav`,
                  forest = rowSums(select(., contains("skog")), na.rm = TRUE),
                  semi_urban = rowSums(select(., contains("ploaterad mark")), na.rm = TRUE),
                  agriculture = `Åkermark`
                  )

NMD <- NMD1 %>% filter(!mvm_id %in% NMD2$mvm_id) %>% bind_rows(NMD2)
NMD
```

## High coast


```{r}

HK <- read_csv(file_HK)
HK
```

## Runoff

```{r}
library(mblm)

discharge.trend <- read_csv(file_discharge.trend) %>% mutate(sampling_date = date, q_m3_s = q, spec_q_mm_day = q/area_m2*1000*24*60*60, log_spec_q_mm_day = log10(spec_q_mm_day + 	0.000000000001 ), year = year(sampling_date)) 

discharge_trend <- discharge.trend %>%
  group_by(mvm_id) %>%
  group_split() %>%
  map_dfr(function(df) {
    tryCatch({
      id <- unique(df$mvm_id)
      uncertainty <- unique(df$model_uncertainty_station_corrected)

      # --- Start period (2008–2012) ---
      start_period <- filter(df, year >= 2008, year <= 2012)

      spec_q_start <- mean(start_period$spec_q_mm_day, na.rm = TRUE)
      log_spec_q_start <- if (all(is.na(start_period$log_spec_q_mm_day))) {
  NA_real_
} else {
  mean(start_period$log_spec_q_mm_day, na.rm = TRUE)
}
      q_start <- mean(start_period$q_m3_s, na.rm = TRUE)

      # --- Change period (2013–2024) ---
      annual_stats <- filter(df, year >= 2013, year <= 2024)%>%
      mutate(
        log_spec_q_mm_day = ifelse(is.finite(log_spec_q_mm_day), log_spec_q_mm_day, NA)) %>%
        group_by(year) %>%
        summarise(
          mean_spec_q = mean(spec_q_mm_day, na.rm = TRUE),
          mean_log_spec_q = mean(log_spec_q_mm_day, na.rm = TRUE),
          mean_q = mean(q_m3_s, na.rm = TRUE),
          .groups = "drop"
        )

      spec_q_change <- if (nrow(annual_stats) >= 2) coef(mblm(mean_spec_q ~ year, data = annual_stats))[2] else NA
      log_spec_q_change <- tryCatch({
            if (nrow(annual_stats) >= 2) {
              coef(mblm(mean_log_spec_q ~ year, data = annual_stats))[2]
            } else {
              NA
            }
          }, error = function(e) NA)
      q_change <- if (nrow(annual_stats) >= 2) coef(mblm(mean_q ~ year, data = annual_stats))[2] else NA

      spec_q_var <- robust_variation(annual_stats$mean_spec_q)
      log_spec_q_var <- robust_variation(annual_stats$mean_log_spec_q)
      q_var <- robust_variation(annual_stats$mean_q)

      tibble(
        mvm_id = id,
        spec_q_start = spec_q_start,
        spec_q_change = spec_q_change,
        spec_q_variation = spec_q_var,
        log_spec_q_start = log_spec_q_start,
        log_spec_q_change = log_spec_q_change,
        log_spec_q_variation = log_spec_q_var,
        q_start = q_start,
        q_change = q_change,
        q_variation = q_var,
        q_uncertainty = uncertainty
      )
    }, error = function(e) {
      message(paste("Error processing mvm_id:", unique(df$mvm_id), "-", e$message))
      return(NULL)
    })
  })

discharge_trend

discharge.omdrev <- read_csv(file_discharge.omdrev) %>% mutate(year = year(date), month = month(date), q_m3_s = q, spec_q_mm_day = q/area_m2*1000*24*60*60, log_spec_q_mm_day = log10(spec_q_mm_day + 	0.000000000001 ))

discharge_sls <- discharge.omdrev %>%
  group_by(mvm_id) %>%
  group_split() %>%
  map_dfr(function(df) {
    tryCatch({
      id <- unique(df$mvm_id)
      uncertainty <- unique(df$model_uncertainty_station_corrected)

      # --- Start period (2008–2012) ---
      start_period <- filter(df, year >= 2008, year <= 2012)

      spec_q_start <- mean(start_period$spec_q_mm_day, na.rm = TRUE)
      log_spec_q_start <- mean(start_period$log_spec_q_mm_day, na.rm = TRUE)
      q_start <- mean(start_period$q_m3_s, na.rm = TRUE)

      # --- Change period (2013–2024) ---
      annual_stats <- filter(df, year >= 2013, year <= 2024)%>%
      mutate(
        log_spec_q_mm_day = ifelse(is.finite(log_spec_q_mm_day), log_spec_q_mm_day, NA)) %>%
        group_by(year) %>%
        summarise(
          mean_spec_q = mean(spec_q_mm_day, na.rm = TRUE),
          mean_log_spec_q = mean(log_spec_q_mm_day, na.rm = TRUE),
          mean_q = mean(q_m3_s, na.rm = TRUE),
          .groups = "drop"
        )

      spec_q_change <- if (nrow(annual_stats) >= 2) coef(mblm(mean_spec_q ~ year, data = annual_stats))[2] else NA
      log_spec_q_change <- tryCatch({
            if (nrow(annual_stats) >= 2) {
              coef(mblm(mean_log_spec_q ~ year, data = annual_stats))[2]
            } else {
              NA
            }
          }, error = function(e) NA)
      q_change <- if (nrow(annual_stats) >= 2) coef(mblm(mean_q ~ year, data = annual_stats))[2] else NA

      spec_q_var <- robust_variation(annual_stats$mean_spec_q)
      log_spec_q_var <- robust_variation(annual_stats$mean_log_spec_q)
      q_var <- robust_variation(annual_stats$mean_q)

      tibble(
        mvm_id = id,
        spec_q_start = spec_q_start,
        spec_q_change = spec_q_change,
        spec_q_variation = spec_q_var,
        log_spec_q_start = log_spec_q_start,
        log_spec_q_change = log_spec_q_change,
        log_spec_q_variation = log_spec_q_var,
        q_start = q_start,
        q_change = q_change,
        q_variation = q_var, 
        q_uncertainty = uncertainty
      )
    }, error = function(e) {
      message(paste("Error processing mvm_id:", unique(df$mvm_id), "-", e$message))
      return(NULL)
    })
  })

discharge <- bind_rows(discharge_trend, discharge_sls) %>%
  distinct(mvm_id, .keep_all = TRUE)

write_csv(discharge, file_discharge.combined)

discharge <- read_csv(file_discharge.combined)

```

## Soil Depth   

```{r}
soil_depth <- read_csv(soil_depth_file)

#soil_depth_corr <- read_csv(r"(C:\Users\anlr0006\repos-win\DOC_catchments\corrections_results\soil_depth.csv)") %>% rename(mvm_id = id)
# soil_depth %>% filter(!(mvm_id %in% soil_depth_corr$mvm_id)) %>%  bind_rows(soil_depth_corr) %>% write.csv(soil_depth_file, row.names = F) 

soil_depth %>% transmute(mvm_id, 
                         soil_depth_mean = mean,
                         soil_depth_IQR = `75th_percentile` - `25th_percentile`
                         ) -> soil_depth
soil_depth
```

## Atmospheric Deposition 

```{r}

emep.1 <- read_csv(emep_file.1) %>% mutate(S_dep_mean_kg_ha_yr = log10(S_dep_mean_kg_ha), N_dep_mean_kg_ha_yr = log10(N_dep_mean_kg_ha))
emep.2 <- read_csv(emep_file.2) %>% mutate(S_dep_mean_kg_ha_yr= log10(S_dep_mean_kg_ha), N_dep_mean_kg_ha_yr = log10(N_dep_mean_kg_ha))

emep.trend <- emep.1  %>%
  group_by(mvm_id) %>%
  group_split() %>%
  map_dfr(function(df) {
    tryCatch({
      id <- unique(df$mvm_id)

      # --- Start period (2008–2012) ---
      start_period <- filter(df, year >= 1990, year <= 1994)

      S_dep_kg_ha_yr_start <- mean(start_period$S_dep_mean_kg_ha_yr, na.rm = TRUE)
      N_dep_kg_ha_yr_start <- mean(start_period$N_dep_mean_kg_ha_yr, na.rm = TRUE)

      # --- Change period (2013–2024) ---
      annual_stats <- filter(df, year >= 1995, year <= 2024)%>%
      mutate(
        S_dep_kg_ha_yr = ifelse(is.finite(S_dep_mean_kg_ha_yr), S_dep_mean_kg_ha_yr, NA),
        N_dep_kg_ha_yr = ifelse(is.finite(N_dep_mean_kg_ha_yr), N_dep_mean_kg_ha_yr, NA)) %>%
        group_by(year) %>%
        
        summarise(
          mean_S_dep_kg_ha_yr = mean(S_dep_kg_ha_yr, na.rm = TRUE),
          mean_N_dep_kg_ha_yr = mean(N_dep_kg_ha_yr, na.rm = TRUE),
          .groups = "drop"
        )


      S_dep_kg_ha_yr_change <- tryCatch({
            if (nrow(annual_stats) >= 2) {
              coef(mblm(mean_S_dep_kg_ha_yr ~ year, data = annual_stats))[2]
            } else {
              NA
            }
          }, error = function(e) NA)

      N_dep_kg_ha_yr_change <- tryCatch({
            if (nrow(annual_stats) >= 2) {
              coef(mblm(mean_N_dep_kg_ha_yr ~ year, data = annual_stats))[2]
            } else {
              NA
            }
          }, error = function(e) NA)

      S_dep_kg_ha_yr_var <- robust_variation(annual_stats$mean_S_dep_kg_ha_yr)
      N_dep_kg_ha_yr_var <- robust_variation(annual_stats$mean_N_dep_kg_ha_yr)

      tibble(
        mvm_id = id,
        S_dep_kg_ha_yr_start = S_dep_kg_ha_yr_start,
        S_dep_kg_ha_yr_change = S_dep_kg_ha_yr_change,
        S_dep_kg_ha_yr_variation = S_dep_kg_ha_yr_var,
        N_dep_kg_ha_yr_start = N_dep_kg_ha_yr_start,
        N_dep_kg_ha_yr_change = N_dep_kg_ha_yr_change,
        N_dep_kg_ha_yr_variation = N_dep_kg_ha_yr_var
      )
    }, error = function(e) {
      message(paste("Error processing mvm_id:", unique(df$mvm_id), "-", e$message))
      return(NULL)
    })
  })

 emep.trend



emep.sls <- emep.2  %>%
  group_by(mvm_id) %>%
  group_split() %>%
  map_dfr(function(df) {
    tryCatch({
      id <- unique(df$mvm_id)

      # --- Start period (1990–1994) ---
      start_period <- filter(df, year >= 1990, year <= 1994)

      S_dep_kg_ha_yr_start <- mean(start_period$S_dep_mean_kg_ha_yr, na.rm = TRUE)
      N_dep_kg_ha_yr_start <- mean(start_period$N_dep_mean_kg_ha_yr, na.rm = TRUE)

      # --- Change period (1995–2024) ---
      annual_stats <- filter(df, year >= 1995, year <= 2024)%>%
      mutate(
        S_dep_kg_ha_yr = ifelse(is.finite(S_dep_mean_kg_ha_yr), S_dep_mean_kg_ha_yr, NA),
        N_dep_kg_ha_yr = ifelse(is.finite(N_dep_mean_kg_ha_yr), N_dep_mean_kg_ha_yr, NA)) %>%
        group_by(year) %>%
        
        summarise(
          mean_S_dep_kg_ha_yr = mean(S_dep_kg_ha_yr, na.rm = TRUE),
          mean_N_dep_kg_ha_yr = mean(N_dep_kg_ha_yr, na.rm = TRUE),
          .groups = "drop"
        )


      S_dep_kg_ha_yr_change <- tryCatch({
            if (nrow(annual_stats) >= 2) {
              coef(mblm(mean_S_dep_kg_ha_yr ~ year, data = annual_stats))[2]
            } else {
              NA
            }
          }, error = function(e) NA)

      N_dep_kg_ha_yr_change <- tryCatch({
            if (nrow(annual_stats) >= 2) {
              coef(mblm(mean_N_dep_kg_ha_yr ~ year, data = annual_stats))[2]
            } else {
              NA
            }
          }, error = function(e) NA)

      S_dep_kg_ha_yr_var <- robust_variation(annual_stats$mean_S_dep_kg_ha_yr)
      N_dep_kg_ha_yr_var <- robust_variation(annual_stats$mean_N_dep_kg_ha_yr)

      tibble(
        mvm_id = id,
        S_dep_kg_ha_yr_start = S_dep_kg_ha_yr_start,
        S_dep_kg_ha_yr_change = S_dep_kg_ha_yr_change,
        S_dep_kg_ha_yr_variation = S_dep_kg_ha_yr_var,
        N_dep_kg_ha_yr_start = N_dep_kg_ha_yr_start,
        N_dep_kg_ha_yr_change = N_dep_kg_ha_yr_change,
        N_dep_kg_ha_yr_variation = N_dep_kg_ha_yr_var
      )
    }, error = function(e) {
      message(paste("Error processing mvm_id:", unique(df$mvm_id), "-", e$message))
      return(NULL)
    })
  })

emep.sls
```

## Chemistry as spatial variable

trend
```{r}
library(dplyr)
library(purrr)
library(lubridate)
library(mblm)

chem.trend <- df.chem.trend %>%
  mutate(year = year(sampling_date)) %>%
  filter(month(sampling_date) %in% c(9,10,11)) %>%
  mutate(
    toc_mg_l_c = log10(toc_mg_l_c),
    totn = log10(totn),
    totp = log10(totp),
    BC_mekv_l = log10(BC_mekv_l),
    DIN_mg_l_n = log10(DIN_mg_l_n),
    cl_mg_l = log10(cl_mg_l),
    so4_mol = log10(so4_mol)
  ) %>%
  group_by(mvp_id) %>%
  group_split() %>%
  map_dfr(function(df) {
    tryCatch({
      id <- unique(df$mvm_id)

      # --- Start period (2008–2012) ---
      start_period <- filter(df, year >= 2008, year <= 2012)

      toc_start   <- mean(start_period$toc_mg_l_c, na.rm = TRUE)
      totn_start  <- mean(start_period$totn, na.rm = TRUE)
      totp_start  <- mean(start_period$totp, na.rm = TRUE)
      ANC_start   <- mean(start_period$ANC_mekv_l, na.rm = TRUE)
      BC_start    <- mean(start_period$BC_mekv_l, na.rm = TRUE)
      DIN_start   <- mean(start_period$DIN_mg_l_n, na.rm = TRUE)
      alk_acid_start <- mean(start_period$alk_acid_mmol, na.rm = TRUE)
      absf_start  <- mean(start_period$absf, na.rm = TRUE)
      pH_start    <- mean(start_period$pH, na.rm = TRUE)
      ec20_start  <- mean(start_period$ec20_µS_cm, na.rm = TRUE)
      cl_start    <- mean(start_period$cl_mg_l, na.rm = TRUE)
      so4_start   <- mean(start_period$so4_mol, na.rm = TRUE)

      # --- Change period (2013–2024) ---
      annual_stats <- filter(df, year >= 2013, year <= 2024) %>%
        group_by(year) %>%
        summarise(
          mean_toc       = mean(toc_mg_l_c, na.rm = TRUE),
          mean_totn      = mean(totn, na.rm = TRUE),
          mean_totp      = mean(totp, na.rm = TRUE),
          mean_ANC       = mean(ANC_mekv_l, na.rm = TRUE),
          mean_BC        = mean(BC_mekv_l, na.rm = TRUE),
          mean_DIN       = mean(DIN_mg_l_n, na.rm = TRUE),
          mean_cl        = mean(cl_mg_l, na.rm = TRUE),
          mean_so4       = mean(so4_mol, na.rm = TRUE),
          mean_alk_acid  = mean(alk_acid_mmol, na.rm = TRUE),
          mean_absf      = mean(absf, na.rm = TRUE),
          mean_pH        = mean(pH, na.rm = TRUE),
          mean_ec20      = mean(ec20_µS_cm, na.rm = TRUE),
          .groups = "drop"
        )


      # Theil–Sen slopes for changes

      toc_change       <- slope_or_na("mean_toc", annual_stats)
      totn_change      <- slope_or_na("mean_totn", annual_stats)
      totp_change      <- slope_or_na("mean_totp", annual_stats)
      ANC_change       <- slope_or_na("mean_ANC", annual_stats)
      BC_change        <- slope_or_na("mean_BC", annual_stats)
      DIN_change       <- slope_or_na("mean_DIN", annual_stats)
      cl_change        <- slope_or_na("mean_cl", annual_stats)
      so4_change       <- slope_or_na("mean_so4", annual_stats)
      alk_acid_change  <- slope_or_na("mean_alk_acid", annual_stats)
      absf_change      <- slope_or_na("mean_absf", annual_stats)
      pH_change        <- slope_or_na("mean_pH", annual_stats)
      ec20_change      <- slope_or_na("mean_ec20", annual_stats)
      
      # Variation calculations
      toc_var       <- robust_variation(annual_stats$mean_toc)
      totn_var      <- robust_variation(annual_stats$mean_totn)
      totp_var      <- robust_variation(annual_stats$mean_totp)
      ANC_var       <- robust_variation(annual_stats$mean_ANC)
      BC_var        <- robust_variation(annual_stats$mean_BC)
      DIN_var       <- robust_variation(annual_stats$mean_DIN)
      cl_var        <- robust_variation(annual_stats$mean_cl)
      so4_var       <- robust_variation(annual_stats$mean_so4)
      alk_acid_var  <- robust_variation(annual_stats$mean_alk_acid)
      absf_var      <- robust_variation(annual_stats$mean_absf)
      pH_var        <- robust_variation(annual_stats$mean_pH)
      ec20_var      <- robust_variation(annual_stats$mean_ec20)


      tibble(
        mvm_id = id,
        toc_start, toc_change, toc_var,
        totn_start, totn_change, totn_var,
        totp_start, totp_change, totp_var,
        ANC_start, ANC_change, ANC_var,
        BC_start, BC_change, BC_var,
        DIN_start, DIN_change, DIN_var,
        cl_start, cl_change, cl_var,
        so4_start, so4_change, so4_var,
        alk_acid_start, alk_acid_change, alk_acid_var,
        absf_start, absf_change, absf_var,
        pH_start, pH_change, pH_var,
        ec20_start, ec20_change, ec20_var
      )

      

    }, error = function(e) {
      message(paste("Error processing mvm_id:", unique(df$mvm_id), "-", e$message))
      return(NULL)
    })
  })


```


```{r}
chem.sls <- df.chem.sls  %>%
  # filter(mvm_id %in% c(1568,1578,2019, 2029, 27051, 40949,39865, 40370))%>% 
  mutate(year = year(sampling_date)) %>%
  filter(month(sampling_date) %in% c(8,9,10,11,12,1)) %>%
  mutate(
    toc_mg_l_c = log10(toc_mg_l_c),
    totn = log10(totn),
    totp = log10(totp),
    BC_mekv_l = log10(BC_mekv_l),
    DIN_mg_l_n = log10(DIN_mg_l_n),
    cl_mg_l = log10(cl_mg_l),
    so4_mol = log10(so4_mol),
    year = year(sampling_date)) %>%
  group_by(mvp_id) %>%
  group_split() %>%
  map_dfr(function(df) {
    tryCatch({
      id <- unique(df$mvp_id)

      # --- Start period (2008–2012) ---
      start_period <- filter(df, year >= 2008, year <= 2012)

      toc_start        <- get_mean_or_first_pre2018(df, "toc_mg_l_c")
      totn_start       <- get_mean_or_first_pre2018(df, "totn")
      totp_start       <- get_mean_or_first_pre2018(df, "totp")
      ANC_start        <- get_mean_or_first_pre2018(df, "ANC_mekv_l")
      BC_start         <- get_mean_or_first_pre2018(df, "BC_mekv_l")
      DIN_start        <- get_mean_or_first_pre2018(df, "DIN_mg_l_n")
      alk_acid_start   <- get_mean_or_first_pre2018(df, "alk_acid_mmol")
      absf_start       <- get_mean_or_first_pre2018(df, "absf")
      pH_start         <- get_mean_or_first_pre2018(df, "pH")
      ec20_start       <- get_mean_or_first_pre2018(df, "ec20_µS_cm")
      cl_start         <- get_mean_or_first_pre2018(df, "cl_mg_l")
      so4_start        <- get_mean_or_first_pre2018(df, "so4_mol")




      

      toc_change        <- get_variable_change(df, "toc_mg_l_c")
      totn_change       <- get_variable_change(df, "totn")
      totp_change       <- get_variable_change(df, "totp")
      ANC_change        <- get_variable_change(df, "ANC_mekv_l")
      BC_change         <- get_variable_change(df, "BC_mekv_l")
      DIN_change        <- get_variable_change(df, "DIN_mg_l_n")
      alk_acid_change   <- get_variable_change(df, "alk_acid_mmol")
      absf_change       <- get_variable_change(df, "absf")
      pH_change         <- get_variable_change(df, "pH")
      ec20_change       <- get_variable_change(df, "ec20_µS_cm")
      cl_change         <- get_variable_change(df, "cl_mg_l")
      so4_change        <- get_variable_change(df, "so4_mol")


        tibble(
  mvm_id = id,
  toc_start, toc_change,
  totn_start, totn_change,
  totp_start, totp_change,
  ANC_start, ANC_change,
  BC_start, BC_change,
  DIN_start, DIN_change,
  cl_start, cl_change,
  so4_start, so4_change,
  alk_acid_start,alk_acid_change,
  absf_start,absf_change,
  pH_start,pH_change,
  ec20_start, ec20_change
)
    }, error = function(e) {
      message(paste("Error processing mvm_id:", unique(df$mvm_id), "-", e$message))
      return(NULL)
    })
  })

library(vtable)
chem.sls %>% sumtable()

chem.sls %>%
  filter(if_any(everything(), is.na))

df.chem.sls %>% filter(mvm_id ==107)
```
  
## NDVI

Error processing ID: 35457 - Must group by variables found in `.data`.

```{r}

file_list <- list.files(path = folder_NDVI, pattern = "*\\.csv$", full.names = FALSE)

file_ids <- gsub("^NDVI_(\\d+)\\.csv$", "\\1", file_list)
missing_ids <- setdiff(id.all, file_ids)



# --- Safely process each ID ---
NDVI.all <- map_dfr(seq_along(id.all), function(i) {
  id <- id.all[i]
  message(sprintf("Processing ID %s (%d of %d)", id, i, length(id.all)))
  
  tryCatch({
    csv.path <- paste0(folder_NDVI,"\\NDVI_", id, ".csv")
    NDVI <- read_csv(csv.path, show_col_types = FALSE) %>%
      filter(year > 2007, year < 2025) %>%
  mutate(
    sampling_date = as.Date(sprintf("%04d-%02d-01", year, as.integer(month)))
  ) %>%
  group_by(year, mvm_id) %>%
  # summarize(NDVI = max(NDVI_median, na.rm = TRUE), .groups = "drop")
    summarize(
  NDVI = if (all(is.na(NDVI_median))) NA_real_ else max(NDVI_median, na.rm = TRUE),
  .groups = "drop"
)

    # --- Start period ---
    start_period <- filter(NDVI, year >= 2008, year <= 2012)

    ndvi_start <- mean(start_period$NDVI, na.rm = TRUE)


    annual_stats <- NDVI %>% filter(year >= 2013, year <= 2024)

    ndvi_change       <- slope_or_na("NDVI", annual_stats)


    ndvi_var <- robust_variation(annual_stats$NDVI)


    tibble(
      mvm_id =as.numeric( id),
      ndvi_start = ndvi_start,
      ndvi_change = ndvi_change,
      ndvi_variation = ndvi_var,

    )
  }, error = function(e) {
    message(paste("Error processing ID:", id, "-", e$message))
    return(NULL)  # Skip this ID
  })
})


NDVI.all%>%
  mutate(across(everything(), ~ifelse(is.na(.), 0, .))) %>% write_csv(file_NDVI)

NDVI.all <- read_csv(file_NDVI)
NDVI.all
```


## Peat and ditches

```{r}
peat <- read.csv(file_peat) %>% select(-ditch_density_m_m2)
peat
```



## Combine everything


```{r}
drivers.sls <- chem.sls %>% mutate(mvm_id = as.integer(mvm_id)) %>%
  left_join(coordinates,     by = "mvm_id") %>%
  left_join(soil_depth, by ="mvm_id") %>%
  left_join(emep.sls,     by = "mvm_id") %>%
  left_join(discharge,    by = "mvm_id")  %>%
  left_join(HK,           by = "mvm_id") %>%
  left_join(climate.all,  by = "mvm_id") %>%
  left_join(general,      by = "mvm_id") %>%
  left_join(NMD,          by = "mvm_id") %>%
  left_join(NDVI.all%>% mutate(mvm_id = as.numeric(mvm_id)),     by = "mvm_id") %>%
  left_join(peat%>% rename(peat_pct_mineralsoil = peat_pct_Mineraljord, peat_pct_peat = peat_pct_Torv_total), by = "mvm_id") %>% mutate(ditch_density_km_km2 = total_ditch_length/ area_m2 * 1000) %>% select(-c(peat_pct_Vatten, total_ditch_length))

drivers.sls

drivers.trend <- chem.trend %>%
  left_join(coordinates,     by = "mvm_id") %>%
  left_join(emep.trend,     by = "mvm_id") %>%
  left_join(discharge,    by = "mvm_id") %>%
  left_join(soil_depth, by ="mvm_id") %>%
  left_join(HK,           by = "mvm_id") %>%
  left_join(climate.all,  by = "mvm_id") %>%
  left_join(general,      by = "mvm_id") %>%
  left_join(NMD,          by = "mvm_id") %>%
  left_join(NDVI.all%>% mutate(mvm_id = as.numeric(mvm_id)),     by = "mvm_id") %>%
  left_join(peat %>% rename(peat_pct_mineralsoil = peat_pct_Mineraljord, peat_pct_peat = peat_pct_Torv_total), by = "mvm_id") %>% mutate(ditch_density_km_km2 = total_ditch_length/ area_m2 * 1000) %>% select(-c(peat_pct_Vatten, total_ditch_length))

drivers.trend
```


# Drivers

```{r}

cols_to_front <- c("mvm_id", "stationname", "x_e_koord","y_n_koord")
drivers.sls%>%
  select(all_of(cols_to_front), everything()) %>% write.csv(file_drivers.sls)

drivers.trend %>%
  select(all_of(cols_to_front), everything())%>% write.csv(file_drivers.trend)

```





